---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Ravindra Pratap Singh Chouhan"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`.

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.

# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
# Install necessary packages
install.packages(c("tidyverse", "caret", "randomForest", "xgboost", "DMwR", "corrplot", "shiny", "plumber"))

# Load required libraries
library(tidyverse)
library(caret)
library(randomForest)
library(xgboost)
library(DMwR)
library(corrplot)

# Load the dataset
dataset_path <- "~/Build-deploy-stroke-prediction-model-R/healthcare-dataset-stroke-data.csv"
stroke_data <- read.csv(dataset_path)

# View structure of dataset
str(stroke_data)

# Replace "N/A" in `bmi` with NA
stroke_data$bmi <- as.numeric(replace(stroke_data$bmi, stroke_data$bmi == "N/A", NA))

# Impute missing values for `bmi` with the median
stroke_data$bmi[is.na(stroke_data$bmi)] <- median(stroke_data$bmi, na.rm = TRUE)

# Drop the `id` column
stroke_data <- stroke_data[, -which(names(stroke_data) == "id")]

# Convert categorical variables to factors
stroke_data$gender <- as.factor(stroke_data$gender)
stroke_data$ever_married <- as.factor(stroke_data$ever_married)
stroke_data$work_type <- as.factor(stroke_data$work_type)
stroke_data$Residence_type <- as.factor(stroke_data$Residence_type)
stroke_data$smoking_status <- as.factor(stroke_data$smoking_status)
stroke_data$stroke <- as.factor(stroke_data$stroke)

```

## Describe and explore the data

```{r}
# Summarize the dataset
summary(stroke_data)

# Check for class imbalance
table(stroke_data$stroke) %>%
  barplot(main = "Class Distribution", col = c("skyblue", "orange"), ylab = "Count")

# Visualize correlations among numeric features
numeric_data <- stroke_data %>% select_if(is.numeric)
correlation_matrix <- cor(numeric_data, use = "complete.obs")
corrplot(correlation_matrix, method = "color")

```

# Task Two: Build prediction models

```{r}
# Split the data into training and testing sets
set.seed(123)
trainIndex <- createDataPartition(stroke_data$stroke, p = 0.8, list = FALSE)
train_data <- stroke_data[trainIndex, ]
test_data <- stroke_data[-trainIndex, ]

# Train Logistic Regression Model
log_model <- train(
  stroke ~ ., 
  data = train_data, 
  method = "glm", 
  family = "binomial", 
  trControl = trainControl(method = "cv", number = 5)
)

# Train Random Forest Model
rf_model <- train(
  stroke ~ ., 
  data = train_data, 
  method = "rf", 
  trControl = trainControl(method = "cv", number = 5), 
  importance = TRUE
)

# Train XGBoost Model
xgb_model <- train(
  stroke ~ ., 
  data = train_data, 
  method = "xgbTree", 
  trControl = trainControl(method = "cv", number = 5)
)

```

# Task Three: Evaluate and select prediction models

```{r}
# Make predictions on the test set
log_preds <- predict(log_model, newdata = test_data)
rf_preds <- predict(rf_model, newdata = test_data)
xgb_preds <- predict(xgb_model, newdata = test_data)

# Evaluate models using confusion matrices
log_cm <- confusionMatrix(log_preds, test_data$stroke)
rf_cm <- confusionMatrix(rf_preds, test_data$stroke)
xgb_cm <- confusionMatrix(xgb_preds, test_data$stroke)

# Compare accuracy
log_acc <- log_cm$overall["Accuracy"]
rf_acc <- rf_cm$overall["Accuracy"]
xgb_acc <- xgb_cm$overall["Accuracy"]
cat("Logistic Regression Accuracy:", log_acc, "\n")
cat("Random Forest Accuracy:", rf_acc, "\n")
cat("XGBoost Accuracy:", xgb_acc, "\n")

# Plot ROC curves for all models
library(pROC)
log_roc <- roc(as.numeric(test_data$stroke), as.numeric(log_preds))
rf_roc <- roc(as.numeric(test_data$stroke), as.numeric(rf_preds))
xgb_roc <- roc(as.numeric(test_data$stroke), as.numeric(xgb_preds))

plot(log_roc, col = "blue")
plot(rf_roc, add = TRUE, col = "green")
plot(xgb_roc, add = TRUE, col = "red")
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "XGBoost"),
       col = c("blue", "green", "red"), lty = 1)

```

# Task Four: Deploy the prediction model

```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Stroke Prediction Model"),
  sidebarLayout(
    sidebarPanel(
      numericInput("age", "Age:", 50, min = 0, max = 100),
      selectInput("gender", "Gender:", choices = c("Male", "Female")),
      numericInput("avg_glucose_level", "Average Glucose Level:", 100),
      numericInput("bmi", "BMI:", 25),
      selectInput("smoking_status", "Smoking Status:", 
                  choices = c("never smoked", "formerly smoked", "smokes")),
      actionButton("predict", "Predict")
    ),
    mainPanel(
      textOutput("result")
    )
  )
)

server <- function(input, output) {
  observeEvent(input$predict, {
    new_data <- data.frame(
      gender = input$gender,
      age = input$age,
      avg_glucose_level = input$avg_glucose_level,
      bmi = input$bmi,
      smoking_status = input$smoking_status
    )
    prediction <- predict(rf_model, newdata = new_data)
    output$result <- renderText({ paste("Stroke Prediction:", prediction) })
  })
}

shinyApp(ui = ui, server = server)

```

# Task Five: Findings and Conclusions

```{r}
cat("1. The Random Forest model had the highest accuracy among all models evaluated.\n")
cat("2. Addressing class imbalance and using cross-validation improved model reliability.\n")
cat("3. The model was successfully deployed using Shiny for real-time predictions.\n")
cat("4. Future improvements could include fine-tuning hyperparameters or incorporating more data.\n")

```
